### 2021-04-13 manuscripts/owp_fim3_2020/sections/introduction.tex
added to methods

Furthermore, \citeA{garousi2019terrain} noted improvements to mapping efficacy by conditioning monotonically decreasing thalweg elevations, adjusting the Manning's n roughness coefficient, and using higher resolution (3m) Digital Elevation Model's (DEM) derived from light detection and ranging (Lidar).

@@

Catchment boundary issues are noted where the interconnection of catchments are not properly accounted for \cite{zhang2018comparative,mcgehee2016modified}.

>>>>

Furthermore, \citeAgarousi2019terrain} noted improvements to mapping efficacy by conditioning monotonically decreasing thalweg elevations, adjusting the Manning's n roughness coefficient, and using higher resolution (3m) Digital Elevation Model's (DEM) derived from light detection and ranging (Lidar).

@@

Catchment boundary issues are noted where the interconnection of catchments are not properly accounted for \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation,nobre2016hand}.

### 2021-04-13 manuscripts/owp_fim3_2020/sections/materials_and_methods.tex
added to methods

@@

The next step involves conditioning these local minimums along the thalweg to enforce monotonically decreasing thalweg elevations for FIM.
\citeA{garousi2019terrain} proposed an algorithm that traverses stream thalweg pixels in a depth first manner starting with adding all the headwater pixels to a queue. 
The connectivity of the thalweg pixels is defined by the D-8 flow directions further discussed in Section \ref{ssec:flow_direction_and_flat_resolution}.
At every thalweg pixel, the minimum elevation among itself and its upstream contributing thalweg pixels is taken as shown in equation \ref{eq:thalweg_breach},
%
\begin{equation}
\label{eq:thalweg_breach}
\textbf{D}[x] = \min_{y\ drains\ to\ x} {(\ \textbf{D}[x]\ ,\ \textbf{D}[y]\ )}
\end{equation}
%
, in which \textbf{D} represents the array of thalweg adjusted elevations indexed by x and y where by y is upstream of x. 
When a pixel's upstream neighbors are all evaluated, the downstream pixel is added into the queue thus the depth first traversal of the drainage network.
This procedure enforces the location of streams and ensures that thalweg elevations are hydrologically correct which yielded a 7\% improvement in critical success index (CSI) per an evaluation for an event in 2017 on the Malad river \cite{garousi2019terrain}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Deriving FIM Hydrofabric}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Flow Directions and Flats Resolution}
\label{ssec:flow_direction_and_flat_resolution}
%
To facilitate the generation of a connected stream network and its associated catchments from the conditioned DEM, the depression-filled DEM is used to derive connectivity in the form of D-8 flow directions.
D-8 seeks to allocate a drainage direction for every pixel based on the adjacent eight pixel neighborhood with the steepest slope \cite{o1984extraction}.
The horizontal component of slope is defined as one for the 4 neighboring pixels in the main cardinal directions while the intercardinal pixels are designated a horizontal component of $\sqrt{2}$. 
The flow direction is encoded as integers 1 through 8 corresponding with the cardinal direction East as 1 and continuing counter-clockwise to the Southwest direction as 8. 
Flow directions are derived for non-depression filled regions trivially with the above procedure but to define connectivity for every grid cell the remaining flats corresponding to depression filled cells must be resolved.

Flat resolution from flats endemic to the DEM or from depression filled regions is a costly,non-trivial procedure which was originally addressed by \citeA{garbrecht1997assignment}.  
Software implementations have developed means to partition the problem and resolve flats iteratively with communication across processes \cite{tarboton2009generalized,tesfa2011extraction,wallis2009parallel,tarboton2005terrain}.
The excessive iteration and communication leads to poor computational performance which motivated further work on how to optimize flat resolution \cite{survila2016scalable,barnes2014efficient}.
Specifically the work by \citeA{survila2016scalable} enables the use of parallel processing and made smoother catchments from our informal experience than those from \citeA{barnes2014efficient}.
By processing flats local to each partition separately from flats shared with other partitions, the accelerated flat resolution algorithm demonstrated an average speed up 468x when compared to prior implementations \cite{survila2016scalable}.
OWP FIM 3 utilized a CyberGIS implementation of the D-8 flow direction algorithm with the accelerated resolution of flats \cite{survila2016scalable,cybergis2016}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Deriving FIM Stream Network}
%
\note[Fernando Aristizabal]{For Brian Avant}
The derivations of relative elevations and catchments from the newly conditioned DEM involves re-deriving a new FIM stream network. 
The FIM stream network is of the same density as the NWM V2.1 network and fully converges at all junctions leaving no divergences in the network.
This is accomplished by using the seed points generated from the stream network enforcement process (Section \ref{ssec:stream_network_enforcment}).
These seeds points are headwater locations of the NHDPlusHR Beta Burnline Events layer that spatially correspond to theheadwater definitions in the stream network of the NWM V2.1.
Feeding the seed points and previously computed flow directions into flow accumulation methods \cite{wallis2009parallel,tarboton1997new,tarboton2005terrain} yields a stream link accumulation raster that can be converted to a vector file for further processing.
Each stream link in this derived FIM stream network is split into equidistant reaches.
Stream links are defined here as segments of rivers discretized by junctions with other NWM river segments.
Stream links are then further segmented at NWM lakes and HUC8 boundaries.
Discretizing at NWM lakes isolates reaches and catchments associated with lakes and reservoirs to avoid mapping them using the Manning's equation and could potentially enable volume based mapping in the future as a feature enhancement.
Additionally every reach (and later catchment) is assigned a globally unique identifier based on the HUC 8 membership.

Now that the stream network has been re-derived using the hydro-conditioning techniques previously discussed, the split stream reaches need to cross-walked to the forecast stream network which is that of the National Water Model's.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Catchments}
%

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Relative Elevation Model}
%

%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Synthetic Rating Curves}
%

>>>>

\label{sssec:stream_thalweg_elevation_conditioning}

@@

The next step involves conditioning these local minimums along the thalweg to enforce monotonically decreasing thalweg elevations for FIM.
\citeA{garousi2019terrain} proposed an algorithm that traverses stream thalweg pixels in a depth first manner starting with adding all the headwater pixels to a queue. 
The connectivity of the thalweg pixels is defined by the D-8 flow directions further discussed in Section \ref{ssec:flow_direction_and_flat_resolution}.
At every thalweg pixel, the minimum elevation among itself and its upstream contributing thalweg pixels is taken as shown in equation \ref{eq:thalweg_breach},
%
\begin{equation}
\label{eq:thalweg_breach}
\textbf{D}[x] = \min_{y\ drains\ to\ x} {(\ \textbf{D}[x]\ ,\ \textbf{D}[y]\ )}
\end{equation}
%
, in which \textbf{D} represents the array of thalweg adjusted elevations indexed by x and y where by y is upstream of x. 
When a pixel's upstream neighbors are all evaluated, the downstream pixel is added into the queue thus the depth first traversal of the drainage network.
This procedure enforces the location of streams and ensures that thalweg elevations are hydrologically correct which yielded a 7\% improvement in critical success index (CSI) per an evaluation for an event in 2017 on the Malad river \cite{garousi2019terrain}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Deriving FIM Hydrofabric}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
The FIM Hydrofabric is defined here as the collection of geospatial datasets that are used for converting NWM discharges into inundation extents.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Flow Directions and Flats Resolution}
\label{ssec:flow_direction_and_flat_resolution}
%
To facilitate the generation of a connected stream network and its associated catchments from the conditioned DEM, the depression-filled DEM is used to derive connectivity in the form of D-8 flow directions.
D-8 seeks to allocate a drainage direction for every pixel based on the adjacent eight pixel neighborhood with the steepest slope \cite{o1984extraction}.
The horizontal component of slope is defined as one for the 4 neighboring pixels in the main cardinal directions while the intercardinal pixels are designated a horizontal component of $\sqrt{2}$. 
The flow direction is encoded as integers 1 through 8 corresponding with the cardinal direction East as 1 and continuing counter-clockwise to the Southwest direction as 8. 
Flow directions are derived for non-depression filled regions trivially with the above procedure but to define connectivity for every grid cell the remaining flats corresponding to depression filled cells must be resolved.

Flat resolution from flats endemic to the DEM or from depression filled regions is a costly,non-trivial procedure which was originally addressed by \citeA{garbrecht1997assignment}.  
Software implementations have developed means to partition the problem and resolve flats iteratively with communication across processes \cite{tarboton2009generalized,tesfa2011extraction,wallis2009parallel,tarboton2005terrain}.
The excessive iteration and communication leads to poor computational performance which motivated further work on how to optimize flat resolution \cite{survila2016scalable,barnes2014efficient}.
Specifically the work by \citeA{survila2016scalable} enables the use of parallel processing and made smoother catchments from our informal experience than those from \citeA{barnes2014efficient}.
By processing flats local to each partition separately from flats shared with other partitions, the accelerated flat resolution algorithm demonstrated an average speed up 468x when compared to prior implementations \cite{survila2016scalable}.
OWP FIM 3 utilized a CyberGIS implementation of the D-8 flow direction algorithm with the accelerated resolution of flats \cite{survila2016scalable,cybergis2016}.
\subsubsection{Deriving FIM Stream Network}
\label{sssec:deriving_fim_stream_network}
\note[Fernando Aristizabal]{For Brian Avant}
The derivations of relative elevations and catchments from the newly conditioned DEM involves re-deriving a new FIM stream network. 
The FIM stream network is of the same density as the NWM V2.1 network and fully converges at all junctions leaving no divergences in the network.
This is accomplished by using the seed points generated from the stream network enforcement process (Section \ref{ssec:stream_network_enforcment}).
These seeds points are headwater locations of the NHDPlusHR Beta Burnline Events layer that spatially correspond to theheadwater definitions in the stream network of the NWM V2.1.
Feeding the seed points and previously computed flow directions into flow accumulation methods \cite{wallis2009parallel,tarboton1997new,tarboton2005terrain} yields a stream link accumulation raster that can be converted to a vector file for further processing.
Each stream link in this derived FIM stream network is split into equidistant reaches.
Stream links are defined here as segments of rivers discretized by junctions with other NWM river segments.
Stream links are then further segmented at NWM lakes and HUC8 boundaries.
Discretizing at NWM lakes isolates reaches and catchments associated with lakes and reservoirs to avoid mapping them using the Manning's equation and could potentially enable volume based mapping in the future as a feature enhancement.
Based on previous research, splitting each remaining stream link into equi-distant reaches not to exceed a parameterized value of 1.5km helps improve synthetic rating curve and mapping skill \cite{garousi2019terrain,godbout2019error,zheng2018geoflood}.
Small reaches can lead to unrealistic variances in channel geometries while too large reaches can lead to grouping too much geometry variance into one parametrized unit.
Additionally every reach (and later catchment) is assigned a globally unique identifier based on the HUC 8 membership.
This stream network is important since it drives the HAND calculation and derivation of catchments.
\subsubsection{Catchments}
Catchments were derived using the D8 connectivity established by \citeA{o1984extraction}.
Gages set at the pixel center points of the delineated stream lines explained in section \ref{sssec:deriving_fim_stream_network}.
The gages act as root nodes in a tree structure and the connectivity is traversed to derive the contributing area for each gage.
Two sets of catchments are derived, one of which assigns the contributing area for each thalweg pixel which is used for relative elevation calculation.
The other catchments is derived for the contributing area for each stream reach as defined in section \ref{sssec:deriving_fim_stream_network}. 
\subsubsection{Height Above Nearest Drainage}
Once the pixel level catchments are derived the final relative elevations can be computed.
Every non-thalweg elevation is substracted from the thalweg elevation within the same pixel-level catchment.
The DEM used for this operation is the DEM resulting from the thalweg conditioning procedures described in section \ref{sssec:stream_thalweg_elevation_conditioning}.
Outside of the excavated channel from the AGREE DEM method, the native non-drainage enforced elevations are used to reduce sources of error in relative elevations due to pit filling. 
\subsubsection{Synthetic Rating Curves}
